Dynamic shifts in predator diel activity patterns across landscapes and threat levels

### 2023-11-01 old_code/manuscript/manuscript.Rmd
store manuscript rmd files

---
title: Spatial variation in predator diel activity patterns
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM,CSIRO
    correspondingauthor: true
  - name: Brendan A. Wintle
    affiliation: UOM
  - name: Jack H. Pascoe
    affiliation: UOM,CEC
  - name: Mark Le Pla
    affiliation: UOM,CEC
  - name: Emma K. Birnbaum
    affiliation: CEC
  - name: Bronwyn A. Hradsky
    affiliation: UOM
address:
  - code: UOM
    address: Quantitative and Applied Ecology Group, School of Ecosystem and Forest Sciences, The University of Melbourne, Parkville, VIC, 3010, Australia
  - code: CSIRO
    address: Health and Biosecurity, Commonwealth Scientific and Industrial Research Organisation, Ecosciences Precinct, Dutton Park, QLD, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Lighthouse Rd, Cape Otway, VIC, 3233, Australia
abstract: |
  *Open research:* Data and code will be deposited on Dryad upon acceptance and can currently be found at this link: https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators.
keywords: 
  - diel activity patterns 
  - generalised additive model
  - feral cat
  - invasive predator
  - intraguild predator interactions
  - lethal predator control
  - mesopredator release
  - red fox
  - spatiotemporal model
journal: "Oikos; Article"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: "thesis-copy.bib"
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: "ecography.csl"
fontsize: 12pt
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    keep_tex: true
    citation_package: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  error     = FALSE,
  cache     = FALSE,
  fig.path  = "../figs/",
  fig.align = "center",
  out.width = "80%")

@@

# Spatial variation in predator diel activity patterns {.unnumbered}

@@

Understanding the constraints dominant predators impose on subordinate species is important for predicting ecosystem dynamics and anticipating outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify antipredator behaviours unless joint spatiotemporal analyses are used.
Here, we tested whether an invasive dominant predator (red fox *Vulpes vulpes*) alters the spatiotemporal activity of an invasive subordinate predator (feral cat *Felis catus*). We collated records of both species from 3,667 camera-traps deployed experimentally across two regions of south-eastern Australia with simplified predator guilds; foxes were poison-baited in some landscapes within each region. We used generalised additive models to quantify changes in predator spatiotemporal activity across geographic space, vegetation types, human footprint and (artificially manipulated) gradients of dominant predator activity.
Foxes and cats had similar diel activity patterns when averaged across all sites, however there was important differentiation at a finer scale---cats did not reduce their spatial activity but shifted diel patterns when localised fox activity was high. Cats were crepuscular on average. However, across dry vegetation types of both regions (where foxes were nocturnal), cats shifted to diurnal behaviour with increasing fox activity. In contrast, fox activity was relatively consistent throughout the daily cycle in the wet forest; here cats avoided dawn when fox activity was high. Changes in cat diel activity patterns may facilitate spatial coexistence between these two invasive predators, potentially shifting feral cat impacts onto different native prey.
It is well-appreciated that predator activity varies spatially and fluctuates throughout the daily cycle. However, our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape-context and fear. dominant predator avoidance appears to be dynamic---a key nuance which is overlooked when simply comparing the average activity overlap between two species.

@@

# INTRODUCTION

Top predators can drive ecosystem structure and dynamics by suppressing populations of rival predators and prey species [@estes2011trophic; @prugh2009rise]. Dominant predators can suppress subordinate species directly through antagonistic interactions and predation, as well as indirectly through fear [@creel2008relationships; @allen2022fear]. Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself, by limiting resource acquisition and breeding success [@schmitz2004trophic; @preisser2005scared; @brown1999ecology]. Understanding how top predators&mdash;including humans&mdash;constrain the behaviour of subordinate species is therefore important to accurately predict consequences of predator management, such as reintroductions or lethal control [@suraci2016fear; @gaynor2021applied].

Spatial and/or temporal niche partitioning may allow subordinate species to coexist with top predators by reducing encounter-rates and resource overlap [@kronfeld2003partitioning]. However, subordinate species may not consistently employ avoidance behaviours because perceived predation risk is spatiotemporally variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times [@lima1999temporal; @lima1990behavioral]. Therefore, optimal predator avoidance strategies likely vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ [@kauffman2007landscape; @willems2009predator; @wirsing2021context; @davies2021spatial]. For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa. These theories are unified under the ecology of fear concept [@brown1999ecology], which has gained increasing attention in recent times [@gaynor2019landscapes]. However, there remains a lack of robust, empirical evidence documenting dynamic changes in antipredator behaviour.

Animal diel activity patterns are increasingly being considered alongside spatial activity and distributional patterns. However, spatial and temporal (in regard to the daily cycle) activity patterns are still often considered in an ad-hoc fashion. For example, by fitting separate models for spatial and temporal overlap, or by repeating spatial analyses (e.g., resource selection functions) at different time periods such night and day [e.g., @basille2015plastic; @kohl2019prey; @smith2019integrating; @wooster2022predator]. Such ad-hoc approaches make it difficult to assess the relative importance, as well as dynamic changes in spatial and temporal processes [@suraci2022beyond].

There are many approaches to modelling intraguild spatiotemporal interactions, such as time-to-encounter models and coefficients of activity overlap [@frey2017investigating]. Time-to-encounter approaches consider fine-scale avoidance or attraction by using time intervals between species detections as the response variable, although do not estimate diel activity patterns [@harmsen2009spatial; @niedballa2019assessing; @fancourt2019introduced]. Multispecies models have been recently developed to assess overlap in spatial and diel activity between two species [@azzou2021sparse], as well as different diel activity patterns based on the presence or absence of another species at a site [@kellner2022two]. Multispecies models are data-hungry and difficult to fit, but can account for imperfect detection and different types of predictor variables [e.g., @parsons2022effect]. Given avoidance behaviours are likely dynamic [@gaynor2019landscapes], there is need for statistical approaches which jointly assess continuous shifts in spatial and diel activity patterns of one species in response to another, particularly when combined with experimental data [@suraci2022beyond].

Generalised Additive Models (hereafter 'GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider spatial activity. GAMs are useful to model diel activity patterns as they can implement cyclical splines&mdash;joining the end with the start of the day&mdash;as well as complexity penalties to reduce overfitting. If complex nonlinear (i.e., wiggly) effects are not statistically supported, smoothing penalties in GAMs 'shrink' them to a linear (i.e., no wiggliness) effect, or even to a horizontal line - removing the entire effect of the predictor variable and therefore  the need for additional tests of statistical significance (such as p-values and information criteria). GAMS can also capture nonlinear interactions between multiple variables with different units (i.e., 'tensor products'), and offer the ability to share information across categorical variables through hierarchical specifications [@wood2017generalized; @pedersen2019hierarchical]. However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM [although without considering overall activity, @cunningham2019temporal].

The red fox *Vulpes vulpes* (hereafter 'fox') and feral cat *Felis catus* (hereafter 'cat') have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \~10 and 63 species, respectively [@doherty2016invasive]. The impacts of these invasive predators are particularly extreme on the Australian continent [@woinarski2015ongoing]. Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes [particularly through poison-baiting, @reddiex2007control]. As foxes and cats compete for many of the same resources [@fleming2022distinctive], there is concern that lethal fox control could cause mesopredator release [@soule1988reconstructed] of cats [@glen2005complex; @doherty2017stop; @wayne2017recoveries]. However, there is limited evidence that spatial cat activity increases in response to fox control [@hunter2018not], although there are signs of changes in behaviour and population density [@molsher2017mesopredator; @rees2023mesopredator]. Other studies have investigated potential spatial and temporal interactions between these invasive predators [e.g., @roshier2021space], but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours with respect to differences in fox activity. Conservation managers are therefore uncertain as to whether fox control may increase the impacts of feral cats on native prey, through increases in cat density or behavioural changes which improve hunting success.

@@

## Study area and camera-trapping
Our study was conducted in a simple predator system where foxes and cats are the only medium-large (> 1 kg) mammalian predators, and fox activity is manipulated using lethal control in some landscapes. This situation allowed sole focus on the interactions between these two predators, across an experimental gradient of dominant predator (fox) activity. We compiled data from multiple smaller-scale camera-trap studies, each designed to experimentally assess mammal responses to fox control. Overall, we collated 5,449 and 2,202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Table \@ref(tab:diel-tab1)).

We collated camera-trap data across two regions in south-west Victoria, Australia: the Glenelg region and Otway Ranges (Fig. \@ref(fig:diel-map)). Here, dingoes *Canis familiaris* are long-absent throughout, while tiger quolls *Dasyurus maculatus* are long-absent in the Glenelg region and likely functionally extinct in the Otway Ranges (last confirmed sighting in 2014). In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation (see sections below for region-specific details). Poison-baits containing 3 mg of sodium fluroacetate (compound 1080) are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads. Different road densities therefore result in variable densities of poison-baits. Foxes readily dig these poison-baits often resulting in population suppression, whereas it is generally accepted that cats are unlikely to consume buried fox-baits and there is no evidence buried 1080 fox-baits suppresses cat populations [@algar2004feral; @moseby2011use; @risbey1997control; @hunter2018not]. Managers also frequently implement prescribed fire across both regions, primarily to reduce fuel loads to prevent large wildfires.

@@

### Otway Ranges
The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig. \@ref(fig:diel-map)). A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east. Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the 'Otway Ark' program [@robley2019otway]. For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement. There was a pause in baiting for approximately six months during the second half of 2018. Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement. A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site [@robley2019otway]. Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one 'before' baiting survey and two 'after' baiting surveys of each site from 2016 - 2018, totalling 1,113 camera-trap deployments [@robley2019otway]. We also include data from an additional before-after control-impact surveys (one 'before' baiting survey and two 'after' baiting surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 [@rees2023data]. This added a further 195 sites and 524 camera-trap deployments (Table \@ref(tab:tab-sum)).

### Camera-trap set-ups
All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure. All camera-trap sites were positioned in the forest interior at least 30 m away from roads. The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away [@robley2020glenelg; @robley2019otway]. The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. The M.W.R PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away [detailed in @rees2023mesopredator]. Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights. Camera-trap spacing averaged 867 m for Glenelg Ark, 1246 m for Otway Ark, and 444 m for the M.W.R PhD surveys (Table \@ref(tab:tab-sum)).

@@

## Generalised additive models 
We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the 'mgcv' R-package [@wood2017generalized]. We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the 'DHARMa' R-package [@DHARMa]. We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling for models 2 and 3 (as pseudoreplication from repeat sampling in model 1 was explicitly accounted for using the spatial term). For fox models, we also included a smooth effect of poison-bait density [number of baits within a 2.3 km radius around each camera-trap - the average maximum distance adult foxes travel from their home-range centre in these regions, @hradsky2017human] with separate responses per region to account for the effect of fox control. These specifications were included in each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the three questions of interest - this is detailed in the sections below. We used the 'ggplot2' [@ggplot2] and 'gratia' [@gratia] R-packages to plot models.

In this paper, we use the term 'spatial activity' to refer to the number of 'independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and 'diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle. The first two models we fit explored how predator spatiotemporal activity varied across these regions, informing the best structure for our third model which tested the activity of foxes limits the spatiotemporal activity of cats (detailed in the sections below).

@@

### How does vegetation type and human footprint affect predator activity patterns? (model 2)
The spatial term in model 1 estimated differences in predator activity across geographic space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions [@swan2015predicting; @hradsky2017bayesian]. To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group [hereafter 'vegetation type' - standard units for vegetation classification in Victoria, @delwp2020bioregions] for each unique camera-trap site, totalling eight vegetation types. As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as 'wet forests'). We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types ['model GS' detailed in @pedersen2019hierarchical]. We also fit a tensor product interaction term for hour and the human footprint index [@venter2016sixteen; @venter2018last]; as human footprint is known to impact animal activity patterns [@gaynor2018influence; @van2023influence]

### Do feral cats avoid foxes in space or time? (model 3)
Fox diel activity across vegetation types showed strong similarity between all vegetation types except wet forests. To examine whether cats avoid foxes in space or time, we therefore modelled fox-induced changes in feral cat diel activity separately for wet forest and dry vegetation types. We further split dry vegetation types by region for further replication. We refer to the resulting variable as 'habitat type', which had three levels: (i) wet forests and rainforests in the western Otway Ranges ('wet_otways'), (ii) dry vegetation types in the Otway Ranges ('dry_otways') and (iii) dry vegetation types in the Glenelg region ('dry_glenelg'). We further hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where fox activity showed little variation across the daily cycle.

To investigate changes in feral cat diel activity across the range of observed fox activity, we first quantified fox activity for each camera trap deployment as the total number of fox detections for the deployment divided by the number of survey days, to adjust for differences in survey duration (hereafter 'adjusted fox counts'). We modelled an interaction between hour and adjusted fox counts (as a thin plate regression spline with shrinkage - so complexity penalties could entirely remove this effect from the model if not supported), allowing cats to have nonlinear responses to both hour and adjusted fox counts. We fit separate tensor product interactions for each habitat type (using a 'by-variable' term). For a direct visual comparison to fox activity, we fit another fox model where a diel activity curve was estimated separately across each of the three habitat types.

@@

>>>>

---
title: Dynamic shifts in predator diel activity patterns across landscapes and threat levels
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM,CSIRO
    correspondingauthor: true
  - name: Brendan A. Wintle
    affiliation: UOM
  - name: Jack H. Pascoe
    affiliation: UOM,CEC
  - name: Mark Le Pla
    affiliation: UOM,CEC
  - name: Emma K. Birnbaum
    affiliation: CEC
  - name: Bronwyn A. Hradsky
    affiliation: UOM
address:
  - code: UOM
    address: Quantitative and Applied Ecology Group, School of Agriculture, Food and Ecosystem Sciences, The University of Melbourne, Parkville, VIC, 3010, Australia
  - code: CSIRO
    address: Health and Biosecurity, Commonwealth Scientific and Industrial Research Organisation, Ecosciences Precinct, Dutton Park, QLD, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Lighthouse Rd, Cape Otway, VIC, 3233, Australia
abstract: |
keywords: 
  - diel activity patterns 
  - generalised additive model
  - feral cat
  - invasive predator
  - intraguild predator interactions
  - lethal predator control
  - mesopredator release
  - red fox
  - spatiotemporal model
journal: "Oikos; Article"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: "thesis-copy.bib"
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: "ecography.csl"
fontsize: 12pt
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    keep_tex: true
    citation_package: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  error     = FALSE,
  cache     = FALSE,
  fig.path  = "",
  fig.align = "center",
  out.width = "80%")

@@

@@

Understanding the constraints dominant predators impose on subordinate species is important for predicting ecosystem dynamics and anticipating outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify antipredator behaviours unless joint spatiotemporal analyses are used.
Here, we tested whether an invasive dominant predator (red fox *Vulpes vulpes*) alters the spatiotemporal activity of an invasive subordinate predator (feral cat *Felis catus*). We collated records of both species from 3,667 camera-traps deployed experimentally across two regions of south-eastern Australia with simplified predator guilds. Foxes were poison-baited in some landscapes within each region. We used generalised additive models to quantify changes in predator spatiotemporal activity across geographic space, vegetation types, human footprint and (artificially manipulated) gradients of dominant predator activity.
Foxes and cats had similar diel activity patterns when averaged across all sites, however there was important differentiation at a finer scale---cats did not reduce their spatial activity but shifted diel patterns when localised fox activity was high. Cats were crepuscular on average. However, across dry vegetation types of both regions (where foxes were nocturnal), cats shifted to diurnal behaviour with increasing fox activity. In contrast, fox activity was relatively consistent throughout the daily cycle in the wet forest; here cats avoided dawn when fox activity was high. Changes in cat diel activity patterns may facilitate spatial coexistence between these two invasive predators, potentially shifting feral cat impacts onto different native prey.
It is well-appreciated that predator activity varies spatially and fluctuates throughout the daily cycle. However, our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape-context and fear. Dominant predator avoidance in time appears to be spatially dynamic---a key nuance which is overlooked when simply comparing the average activity overlap between two species.

@@

# INTRODUCTION
Top predators can drive ecosystem structure and dynamics by suppressing populations of rival predators and prey species [@estes2011trophic; @prugh2009rise]. Dominant predators can suppress subordinate species directly through antagonistic interactions and predation, as well as indirectly through fear [@creel2008relationships]. Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself, by limiting resource acquisition and breeding success [@schmitz2004trophic; @preisser2005scared; @brown1999ecology]. Understanding how top predators&mdash;including humans&mdash;constrain the behaviour of subordinate species is therefore important to accurately predict consequences of predator management, such as reintroductions or lethal control [@gaynor2021applied].

Spatial and/or temporal niche partitioning may allow subordinate species to coexist with top predators by reducing encounter-rates and resource overlap [@kronfeld2003partitioning]. However, subordinate species may not consistently employ avoidance behaviours because perceived predation risk is spatiotemporally variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times [@lima1999temporal; @lima1990behavioral]. Therefore, optimal predator avoidance strategies likely vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ [@kauffman2007landscape; @willems2009predator; @wirsing2021context; @davies2021spatial; @gaynor2018influence]. For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa. These theories are unified under the ecology of fear concept [@brown1999ecology], which has gained increasing attention in recent times [@gaynor2019landscapes]. However, there remains a lack of robust, empirical evidence documenting dynamic changes in antipredator behaviour [@gilbert2022daily].

Animal diel activity patterns are increasingly being considered alongside spatial activity and distributional patterns. However, spatial and temporal (in regard to the daily cycle) activity patterns are still often considered in an ad-hoc fashion. For example, by fitting separate models for spatial and temporal overlap, or by repeating spatial analyses (e.g., resource selection functions) at different time periods such night and day [e.g., @basille2015plastic; @kohl2019prey; @smith2019integrating; @wooster2022predator]. Such approaches make it difficult to assess the relative importance, as well as dynamic changes in spatial and temporal processes.

There are many approaches to modelling intraguild spatiotemporal interactions, such as time-to-encounter models and coefficients of activity overlap [@frey2017investigating]. Time-to-encounter approaches consider fine-scale avoidance or attraction by using time intervals between species detections as the response variable, but do not estimate diel activity patterns [@harmsen2009spatial; @niedballa2019assessing; @fancourt2019introduced]. Multispecies models have been recently developed to assess overlap in spatial and diel activity between two species [@azzou2021sparse], as well as different diel activity patterns based on the presence or absence of another species at a site [@kellner2022two]. Multispecies models are data-hungry and difficult to fit, but can account for imperfect detection and different types of predictor variables [e.g., @parsons2022effect]. Given avoidance behaviours are likely dynamic [@gaynor2019landscapes], there is need for statistical approaches which jointly assess continuous shifts in spatial and diel activity patterns of one species in response to another, particularly when combined with experimental data [@suraci2022beyond].

Generalised Additive Models (hereafter 'GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider spatial activity. GAMs are useful for modelling diel activity patterns as they can implement cyclical splines, which join the end of the one day with the start of the following day. GAMs can also include complexity penalties to reduce overfitting; these penalties can reduce complex nonlinear (i.e., wiggly) smooths to linear effects or even remove the effect from the model if it is not statistically supported, avoiding the need for additional tests of statistical significance (such as p-values and information criteria). GAMS can also capture nonlinear interactions between two variables with different units (i.e., 'tensor products'), and offer the ability to share information across categorical variables through hierarchical specifications [@wood2017generalized; @pedersen2019hierarchical]. However, we are only aware of one other study that has used a GAM to explore how animal diel activity interacts with predation risk as a continuous variable: @cunningham2019temporal used this approach to document shifts in diel activity patterns of mammal communities following decline in a top predator (although without also considering spatial animal activity).

Invasive predators are often lethally controlled to protect native competitor and prey species; however, the conservation benefits of invasive predator control may be dampened if this benefits other invasive species [@takimoto2022simple; @doherty2017stop]. The red fox *Vulpes vulpes* (hereafter 'fox') and feral cat *Felis catus* (hereafter 'cat') have had devastating impacts on native prey throughout their introduced range, and are implicated in the extinction of \~10 and 63 species worldwide, respectively [@doherty2016invasive]. The impacts of these invasive predators are particularly extreme on the Australian continent [@woinarski2015ongoing]. Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes [particularly through poison-baiting, @reddiex2007control]. As foxes and cats compete for many of the same resources [@fleming2022distinctive], there is concern that lethal fox control could cause mesopredator release [@soule1988reconstructed] of cats [@glen2005complex; @wayne2017recoveries]. While there is evidence that foxes may suppress cat density and alter movement behaviour [@molsher2017mesopredator; @rees2023mesopredator], potential changes in cat diel activity have been relatively rarely studied [although see @roshier2021space]. Changes in predator diel activity patterns may lead to changes in population-level fitness, as well as hunting success and primary prey species [@allen2022fear; @suraci2016fear].

@@

## Study area and camera-trapping
Our study was conducted in a simple predator system where foxes and cats are the only medium-large (> 1 kg) mammalian predators, and fox activity is manipulated using lethal control in some landscapes. This situation allowed sole focus on the interactions between these two predators, across an experimental gradient of dominant predator (fox) activity. We compiled data from multiple smaller-scale camera-trap studies, each designed to experimentally assess mammal responses to fox control. Overall, we collated 5,449 and 2,202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Supporting Information Table \@ref(tab:diel-tab1)).

We collated camera-trap data across two regions in south-west Victoria, Australia: the Glenelg region and Otway Ranges (Fig. \@ref(fig:diel-map)). Here, dingoes *Canis familiaris* are long-absent throughout, while tiger quolls *Dasyurus maculatus* are long-absent in the Glenelg region and likely functionally extinct in the Otway Ranges (last confirmed sighting in 2014). In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation (see sections below for region-specific details). Poison-baits containing 3 mg of sodium fluroacetate (compound 1080) are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads. Different road densities therefore result in variable densities of poison-baits. Foxes readily dig these poison-baits, whereas it is generally accepted that cats are unlikely to consume buried fox-baits and there is no evidence that buried 1080 fox-baits suppress cat populations [@algar2004feral; @moseby2011use; @risbey1997control; @hunter2018not]. Managers also frequently implement prescribed fire across both regions, primarily to reduce fuel loads to prevent large wildfires.

@@

### Otway Ranges
The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig. \@ref(fig:diel-map)). A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east. Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the 'Otway Ark' program [@robley2019otway]. For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement. There was a pause in baiting for approximately six months during the second half of 2018. Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement. A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site [@robley2019otway]. Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one 'before' baiting survey and two 'after' baiting surveys of each site from 2016 - 2018, totalling 1,113 camera-trap deployments [@robley2019otway]. We also include data from an additional before-after control-impact surveys (one 'before' baiting survey and two 'after' baiting surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 [@rees2023data]. This added a further 195 sites and 524 camera-trap deployments (Supporting Information Table \@ref(tab:tab-sum)).

### Camera-trap set-ups
All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure. All camera-trap sites were positioned in the forest interior at least 30 m away from roads. The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away [@robley2020glenelg; @robley2019otway]. The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. The M.W.R PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away [detailed in @rees2023mesopredator]. Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights. Camera-trap spacing averaged 867 m for Glenelg Ark, 1246 m for Otway Ark, and 444 m for the M.W.R PhD surveys (Supporting Information Table \@ref(tab:tab-sum)).

@@

## Generalised additive models 
We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the 'mgcv' R-package [@wood2017generalized]. We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the 'DHARMa' R-package [@DHARMa]. We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling for models 2 and 3 (pseudoreplication from repeat sampling in model 1 was explicitly accounted for using the spatial term). For all fox models, we also included a smooth effect of poison-bait density [the number of baits within a 2.3 km radius around each camera-trap - the average maximum distance adult foxes travel from their home-range centre in these regions, @hradsky2017human] with separate responses in each region to account for the effect of fox control effort. Poison-bait density ranged from 0 - 19 baits per 2.3km radius in both the Glenelg and Otway region. Models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the three questions of interest - this is detailed in the sections below. We used the 'ggplot2' [@ggplot2] and 'gratia' [@gratia] R-packages to plot models.

In this paper, we use the term 'spatial activity' to refer to the number of 'independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and 'diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle. The first two models explore how predator spatiotemporal activity varies across these regions. We used these to inform the best structure for our third model which tested whether the activity of foxes limits the spatiotemporal activity of cats (detailed in the sections below).

@@

### How does vegetation type and human footprint affect predator activity patterns? (model 2)
The spatial term in model 1 estimated differences in predator activity across geographic space; we hypothesised that this would be partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions [@swan2015predicting; @hradsky2017bayesian]. We further hypothesised that predators would become relatively more nocturnal with increasing human activity, reflecting temporal avoidance of humans [@gaynor2018influence].

To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group [hereafter 'vegetation type' - standard units for vegetation classification in Victoria, @delwp2020bioregions] for each unique camera-trap site, totalling eight vegetation types. As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as 'wet forests'). We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types ['model GS' detailed in @pedersen2019hierarchical].

To influence of human activity on predator activity patterns, we fit a tensor product interaction term for hour and the human footprint index [@venter2018last]. The human footprint index maps human pressure on environments by considering variables such as human population density, infrastructure and agriculture [@venter2016sixteen].

### Do feral cats avoid foxes in space or time? (model 3)
Model 2 showed that fox diel activity patterns was mostly uniform among all vegetation types except wet forests. To examine whether cats avoid foxes in space or time, we therefore modelled fox-induced changes in feral cat diel activity separately for wet forest and dry vegetation types. We examined responses in dry vegetation types by region for further replication. We refer to the resulting variable as 'habitat type', which had three levels: (i) wet forests and rainforests in the western Otway Ranges ('wet_otways'), (ii) dry vegetation types in the Otway Ranges ('dry_otways') and (iii) dry vegetation types in the Glenelg region ('dry_glenelg'). We hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where foxes were more consistently active throughout the 24-hour period.

To investigate changes in feral cat diel activity across the range of observed fox activity, we first quantified fox activity for each camera trap deployment as the total number of fox detections for the deployment divided by the number of survey days, to adjust for differences in survey duration (hereafter 'adjusted fox counts'). We modelled an interaction between hour and adjusted fox counts (as a thin plate regression spline with shrinkage - complexity penalties could entirely remove this effect from the model if not supported), allowing cats to have nonlinear responses to both hour and adjusted fox counts. We fit separate tensor product interactions for each habitat type (using a 'by-variable' term). For a direct visual comparison to fox activity, we fit another fox model where a diel activity curve was estimated separately across each of the three habitat types.

@@

# ACKNOWLEDGEMENTS {.unnumbered}
